Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices

### 2015-09-20 flux-entropy-paper/flux-entropy-paper.tex
updates to refs so paper will compile

  %Digital arts practices can allow for the archival of interaction log data.
  %Such log data can capture and preserve performances of new-media artists and musicians and
  %and can facilitate post-hoc data processing and data mining.

  %In this note we describe a data archive for an iPad musical-ensemble performance practice which has included 
  %concerts, demonstrations and formal HCI studies of groups of musicians over more than two years. Our corpus contains data for 
  %150 performances, rehearsals, and demonstrations, involving some 100 different musicians and several different iPad apps.
  %At the lowest level, data for each musician in a particular performance was archived as Apple iOS callback methods for touching and swiping. Subsequent data analysis has been undertaken by first classifying these data into higher level gesture types and then describing the transitions between these gestures for each artist as a function of time. We 
%find that a transition-matrix measure, which we call ``flux'', can be
%used to distinguish different performance types and contexts, as well as different phases (beginning, middle and end) of all of the performances in our corpus.

@@

Modern computing devices and interfaces continue to open up rich, open-ended
creative opportunities in the new media arts. 
Understanding and evaluating interfaces in these contexts presents many challenges
[refs...Resnick, Mitchel, Brad Myers, Kumiyo Nakakoji, Ben Shneiderman, Randy Pausch, Ted Selker, and Mike Eisenberg. "Design principles for tools to support creative thinking." (2005) http://repository.emu.edu/isr; Shneiderman, Ben. "Creativity support tools: Accelerating discovery and innovation." Communications of the ACM 50.12 (2007): 20-32.]. 
In the context of live performance, interface evaluation, using, for example, post-performance questionnaires, can omit consideration of the dynamic needs of artists in the moment of the performance and in the wild of the performance context. 
In contrast to traditional art forms, logged interaction data from a new media arts performance can provide an enduring record of that performance for post-hoc quantitative investigation as well as for curation and preservation[ref: England, David, et al. "Curating the digital: Spaces for art and interaction." CHI'14 Extended Abstracts on Human Factors in Computing Systems. ACM, 2014.].
Over time, the accumulation of new-media interaction log data might allow data-mining techniques to be deployed to potentially describe universal features of such interaction. 
%Additionally, a wider dissemination of such log data will better support the replication of findings in HCI research [ref:Wilson, Max L., et al. "RepliCHI-CHI should be replicating and validating results more: discuss." CHI'11 Extended Abstracts on Human Factors in Computing Systems. ACM, 2011.]

It is a feature of creative  expression in the arts that highly skilled practitioners exploit simple tools to produce works of beauty whether it be static paintings or dynamic performance. In previous work [ref: charles at CHI] researchers have shown that trained percussionists can exploit a very simple iPad interface to discover new categories of musical gestures. In that paper, iPad apps were studied where tapping was mapped to produce produced short sounds, while swiping produced long sounds
with the volume proportional to the velocity of the moving touch point. It turned out that trained percussionists working together in an ensemble were able to discover and repeatedly make use of five macro gestures made up of taps and swipes: fast taps, slow taps, random taps, short swipes and swirls. This process of discovery of interesting musical gestures from a simple interface is analogous to what percussionists do with acoustic instruments -- drums and bowls and bells are often tapped, stroked and hit in novel ways to communicate meaning and emotion through music [ref to Schick ].

In this note we describe a log data archive that has been accumulated to 
support a long-term practice-led research program into simple iPad
interfaces for group music-making similar to those described in [ref: charles at CHI] . Over a period of more than two
years (April 2013--June 2015), we collected data from $....$ group
music-making sessions where the participants used custom-built, touch-screen
apps on Apple iPad devices. Although several different apps have been included in our corpus, (see\cite{anon} for
details) they all shared the simple mode of interaction where the majority of
the screen could be used for free-form touch interactions; tapping
produced short sounds, while swiping or swirling produced long sounds
with the volume proportional to the velocity of the moving touch point. Apart from this commonality, our apps featured different palettes of available sounds and modes of synthesis, different arrangements of sounds on the iPad screens, different kinds of visual display and feedback, different kinds of networked interactions (between the iPads in an ensemble and with a central server), as well as a variety of special features for each app.

@@

A useful way of modelling user interfaces is to consider their configurations as state vectors and to deploy the methods of matrix algebra to describe transitions between these states during an interaction sequence\cite{Thimbleby:2001kq, Thimbleby, Harold, Paul Cairns, and Matt Jones. "Usability analysis with Markov models." ACM Transactions on Computer-Human Interaction (TOCHI) 8.2 (2001): 99-132.}. Data obtained from usability analysis of quite complex systems (such as a resource planning application for anti-terrorism planning [ref: Kannampallil, Thomas, and Steven Haynes. "Using Event Based Markov Model Simulation to Analyze Interactive Human Behavior." AMCIS 2007 Proceedings (2007): 142. ]) have been undertaken by coding sequences of fine-detail human actions as states and then to model data from usability walkthroughs as sequences of state transitions. The transition probably matrix for transitions between states can then be derived from such user data and conclusions drawn about the nature of interaction. In the creative arts, editing commands at a computer interface for the ``live coding of computer music'' have also been coded coded as states and analysed using transition matrices, in both the textual and musical dimensions, in order to arrive at conclusions regarding artistic style\cite{Swift:2014tya}.

@@

Mention Manovich transmutation stuff

% todo blah de blah

>>>>

%   %Digital arts practices can allow for the archival of interaction
% log data. Such log data can capture and preserve performances of
% new-media artists and musicians and %and can facilitate post-hoc data
% processing and data mining.

%   %In this note we describe a data archive for an iPad
% musical-ensemble performance practice which has included %concerts,
% demonstrations and formal HCI studies of groups of musicians over more
% than two years. Our corpus contains data for %150 performances,
% rehearsals, and demonstrations, involving some 100 different musicians
% and several different iPad apps. %At the lowest level, data for each
% musician in a particular performance was archived as Apple iOS
% callback methods for touching and swiping. Subsequent data analysis
% has been undertaken by first classifying these data into higher level
% gesture types and then describing the transitions between these
% gestures for each artist as a function of time. We %find that a
% transition-matrix measure, which we call ``flux'', can be %used to
% distinguish different performance types and contexts, as well as
% different phases (beginning, middle and end) of all of the
% performances in our corpus.

@@

Modern computing devices and interfaces continue to open up rich,
open-ended creative opportunities in the new media arts. Understanding
and evaluating interfaces in these contexts presents many challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}. In the context of live performance,
interface evaluation, using, for example, post-performance
questionnaires, can omit consideration of the dynamic needs of artists
in the moment of the performance and in the wild of the performance
context. In contrast to traditional art forms, logged interaction data
from a new media arts performance can provide an enduring record of
that performance for post-hoc quantitative investigation as well as
for curation and preservation~\cite{England:2014ys}. Over
time, the accumulation of new-media interaction log data might allow
data-mining techniques to be deployed to potentially describe
universal features of such interaction.

%Additionally, a wider dissemination of such log data will better support the replication of findings in HCI research [ref:Wilson, Max L., et al. "RepliCHI-CHI should be replicating and validating results more: discuss." CHI'11 Extended Abstracts on Human Factors in Computing Systems. ACM, 2011.]

It is a feature of creative expression in the arts that highly skilled
practitioners exploit simple tools to produce works of beauty whether
it be static paintings or dynamic performance. In previous
work~\cite{Martin:2014cr} researchers have shown that trained
percussionists can exploit a very simple iPad interface to discover
new categories of musical gestures. In that paper, iPad apps were
studied where tapping was mapped to produce produced short sounds,
while swiping produced long sounds with the volume proportional to the
velocity of the moving touch point. It turned out that trained
percussionists working together in an ensemble were able to discover
and repeatedly make use of five macro gestures made up of taps and
swipes: fast taps, slow taps, random taps, short swipes and swirls.
This process of discovery of interesting musical gestures from a
simple interface is analogous to what percussionists do with acoustic
instruments -- drums and bowls and bells are often tapped, stroked and
hit in novel ways to communicate meaning and emotion through music~\cite{Schick:2006fk}.

In this note we describe a log data archive that has been accumulated
to support a long-term practice-led research program into simple iPad
interfaces for group music-making similar to those described in [ref:
charles at CHI] . Over a period of more than two years (April
2013--June 2015), we collected data from $....$ group music-making
sessions where the participants used custom-built, touch-screen apps
on Apple iPad devices. Although several different apps have been
included in our corpus,
% TODO: should there be a citation here?
 they all shared
the simple mode of interaction where the majority of the screen could
be used for free-form touch interactions; tapping produced short
sounds, while swiping or swirling produced long sounds with the volume
proportional to the velocity of the moving touch point. Apart from
this commonality, our apps featured different palettes of available
sounds and modes of synthesis, different arrangements of sounds on the
iPad screens, different kinds of visual display and feedback,
different kinds of networked interactions (between the iPads in an
ensemble and with a central server), as well as a variety of special
features for each app.

@@

A useful way of modelling user interfaces is to consider their configurations as state vectors and to deploy the methods of matrix algebra to describe transitions between these states during an interaction sequence\cite{Thimbleby:2001kq, Thimbleby:2004fj}. Data obtained from usability analysis of quite complex systems (such as a resource planning application for anti-terrorism planning [ref: Kannampallil, Thomas, and Steven Haynes. "Using Event Based Markov Model Simulation to Analyze Interactive Human Behavior." AMCIS 2007 Proceedings (2007): 142. ]) have been undertaken by coding sequences of fine-detail human actions as states and then to model data from usability walkthroughs as sequences of state transitions. The transition probably matrix for transitions between states can then be derived from such user data and conclusions drawn about the nature of interaction. In the creative arts, editing commands at a computer interface for the ``live coding of computer music'' have also been coded coded as states and analysed using transition matrices, in both the textual and musical dimensions, in order to arrive at conclusions regarding artistic style\cite{Swift:2014tya}.

@@

%TODO Mention Manovich transmutation stuff
